---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# validateR

<!-- badges: start -->
<!-- badges: end -->

`validateR` is a lightweight toolkit for checking data frames for common data quality issues: missing values, outliers, duplicate rows, and type inconsistencies. Results are returned as a structured report object with `print` and `plot` methods, so you can quickly see what's wrong with your data before you start analyzing it.

## Installation

You can install the development version of validateR from [GitHub](https://github.com/UgyenNorbu/validateR) with:

``` r
pak::pak("UgyenNorbu/validateR")
```
## Example

The main entry point is `validate_df()`, which runs all four checks at once and returns a single report:

```{r example}
library(validateR)

messy_df <- data.frame(
  age = c(25, 30, NA, 45, 22, 31, 29, 200),
  income = c("50000", "62000", "N/A", "48000",
             "71000", "55000", "60000", "48000"),
  city = c("Thimphu", "Paro", "Thimphu", "Paro",
           "Thimphu", "Paro", "Punakha", "Paro"),
  stringsAsFactors = FALSE
)
messy_df <- rbind(messy_df, messy_df[1, ], messy_df[4, ])

report <- validate_df(messy_df)
report
```

You can drill into the full detail behind any of the four checks:

```{r cars}
report$results$missing
report$results$types
```

And visualize missing values and outliers per column:

```{r report, echo = FALSE}
plot(report)
```

## Individual checks

Each check is also available on its own, if you only need one:

```{r individual}
check_duplicates(messy_df)
```
