---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# validateR

<!-- badges: start -->
<!-- badges: end -->

`validateR` is a lightweight toolkit for checking data frames for common data quality issues: missing values, outliers, duplicate rows, numeric-like inconsistencies, and (optionally) date-like inconsistencies. Results are returned as a structured report object with `print` and `plot` methods, so you can quickly see what's wrong with your data before you start analyzing it.

## Installation

You can install the development version of validateR from [GitHub](https://github.com/UgyenNorbu/validateR) with:

``` r
pak::pak("UgyenNorbu/validateR")
```
## Example

The main entry point is `validate_df()`, which runs several checks at once and returns a single report:

```{r example}
library(validateR)

messy_df <- data.frame(
  age = c(25, 30, NA, 45, 22, 31, 29, 200),
  income = c("50000", "62000", "N/A", "48000",
             "71000", "55000", "60000", "48000"),
  city = c("Thimphu", "Paro", "Thimphu", "Paro",
           "Thimphu", "Paro", "Punakha", "Paro"),
  stringsAsFactors = FALSE
)
messy_df <- rbind(messy_df, messy_df[1, ], messy_df[4, ])

report <- validate_df(messy_df)
report
```

You can drill into the full detail behind any of the checks:

```{r drill-down}
report$results$missing
report$results$numeric_like
```

And visualize missing values and outliers per column:

```{r plot-example, echo = FALSE}
plot(report)
```

## Checking date-like columns

`validate_df()` can also check whether character columns contain values that look like dates in a format you specify. This check is optional, since date formats are ambiguous (for example, `"03/04/2023"` could mean March 4th or April 3rd, depending on convention) — so rather than guessing, you tell it what format to expect for each column you want checked.

```{r date-example}
dated_df <- data.frame(
  signup_date = c("2023-01-15", "2023-02-20", "not a date",
                  "2023-03-10", "2023-04-01"),
  stringsAsFactors = FALSE
)

report_with_dates <- validate_df(
  dated_df,
  date_columns = c(signup_date = "%Y-%m-%d")
)
report_with_dates
```

If `date_columns` isn't supplied, `validate_df()` skips date checking entirely, and the report notes that it wasn't checked rather than implying the column was fine:

```{r no-date-example}
report_without_dates <- validate_df(dated_df)
report_without_dates
```

## Individual checks

Each check is also available on its own, if you only need one:

```{r individual}
check_duplicates(messy_df)
```
